\vspace{1.0cm} \LARGE \bf Reassortment between influenza B lineages and the emergence of a co-adapted PB1-PB2-HA gene complex

### 2013-12-20 fluB.tex
Include railroad track plot in text.
Using the railroad track plot and elaborating on some lineages with
reference to the plot.
Also adding more information with regards to mixed-lineage reassortant
lineages (number of sequences, dates of first and last isolate).

The split of Vic and Yam lineages can be seen in all segments.
Following the split of the two lineages, each segment could be assigned to either Vic or Yam lineage and inter-lineage reassortment events have yielded mixed-lineage genome constellations.
Some viruses with mixed-lineage genomes have become common ancestors to many later strains (i.e. the `trunk' of the phylogenetic tree) and the segment lineages involved so successful, as to become fixed within the influenza B population circulating today.

@@

\subsection*{PB1, PB2 and HA segments continue to circulate as two lineages and have not reassorted recently}
Unlike PA, NP, NA, MP and NS segments, which have periodically lost diversity, we find that PB1, PB2 and HA segments have continued to accumulate diversity since the initial split of Vic and Yam lineages (see Figure \ref{tmrcaOT}).
In addition, there is evidence to suggest that the numbers of sequences that belong to either Vic and Yam lineages of PB1, PB2 and HA segments have been sampled at a ratio close to 0.5 (see Figure \ref{lineageRatiosOverTime}), presumably due to the action of some form of balancing selection, possibly to maintain antigenic diversity.

@@

Here we focus on PB1, PB2 and HA lineage labels and find that mean pairwise diversity between branches labelled as having for example, a PB1 segment of Vic or Yam lineage in PB2 and HA trees (and other combinations of the 3 trees and their labels), show small drops in diversity initially, but from the 1997 time slice onwards all branches labelled as Vic or Yam have a common ancestor at the root of the tree.
We also find evidence of co-reassortment in PB1, PB2 and HA segments: in Figure \ref{betweenDiversity} all three graphs become identical after 1997, indicating that the same PB1, PB2 and HA labels have been assigned to the same branches in trees of all other segments.

@@

\subsection*{Victoria and Yamagata lineages of PB1-PB2-HA are co-adapted segment complexes}

@@

We find that PB1, PB2 and HA segments, in addition to not reassorting across the Vic-Yam lineage boundary recently, form co-reassorting segment complexes derived entirely of Vic or Yam lineage segments.
Figure \ref{stateTime} shows the sum of branch lengths which were labelled as having entirely Vic, entirely Yam or mixed-lineage PB1, PB2 and HA segments.

\subsection*{PB1-PB2-HA reassortants are rare and transient}
Figures \ref{betweenDiversity} and \ref{stateTime} suggests that some inter-lineage reassortment has occured between PB1, PB2 and HA segments soon after the split of Vic and Yam lineages, but the resulting viruses were not fit and did not become the `trunk' of the tree.
From complete genome data we have identified 3 classes of PB1-PB2-HA reassortants with the following PB1-PB2-HA constellations: VVY (B/Bankok/163/1990-like), YVV (B/Nanchang/630/1994-like) and YVY (B/New York/24/1993-like).
An additional class of reassortants with a YYV PB1-PB2-HA constellation (B/Waikato/6/2005-like) were found when investigating a larger dataset with only PB1, PB2, HA and NA sequences.
These reassortants are represented by 16 sequences and were isolated in New Zealand and Australia between June 9 and November 12 in 2005.
In all cases PB1-PB2-HA reassortants have not persisted for prolonged periods of time and have not been fixed in the influenza B population.

\subsection*{Other segments evolve with considerable independence}

The reconstructed evolution of Victoria and Yamagata lineages of all segments reveals a considerable degree of autonomy.
For example, of all the Yamagata lineage segments that were reassorted into Victoria lineage PB1-PB2-HA background giving rise to B/Hong Kong/330/2001-like viruses, most segments went extinct following later reassortments, save for NP, which continued circulating until, at the latest, 2011.
A more dramatic example can be found in the NS segment, where the Vic lineage had been fixed in the population after reassorting into an otherwise entirely Yam-derived genome constellation and then reassorted back into a Vic lineage PB1-PB2-HA background, replacing the NS lineage that was present there.

@@

\subsection*{Reassortment rates are not significantly different between PB1, PB2 and HA, but reassortment distances can be}
Estimates of normalized SPR distances between pairs of segments (Figure \ref{SPRdistances}) show that there is no evidence of differences in reassortment rates between all segments, based on overlapping 95\% highest posterior density (HPD) intervals.
\begin{figure}
	\centering		
	\includegraphics[width=0.95\textwidth]{figures/InfB_normalizedApproxSPR.png}
	\caption{\textbf{Normalized approximate SPR distances between pairs of segments.}
All vertical lines indicating uncertainty are 95\% highest posterior densities (HPDs).}
	\label{SPRdistances}
\end{figure}

Unlike normalized SPR distances normalized $\Delta$TMRCA estimates between some segments (Figure \ref{deltaTMRCA}) indicate that reassortment distance (as quantified by $\Delta$TMRCA) between some segments, namely PB1-PB2-HA (but also NA-M1 and PA-M1-NS1), are much more similar to $\Delta$TMRCA estimates between replicate trees.
\begin{figure}
	\centering		
	\includegraphics[width=0.95\textwidth]{figures/InfB_normalizedMuDeltaTMRCA.png}
	\caption{\textbf{Normalized mean $\Delta$TMRCA statistics between pairs of segments.}
All vertical lines indicating uncertainty are 95\% highest posterior densities (HPDs).}
	\label{deltaTMRCA}
\end{figure}

@@

>>>>

The split of Vic and Yam lineages can be seen in all segments \cite{chen2008}.
Following the split of the two lineages, each segment could be assigned to either Vic or Yam lineage and inter-lineage reassortment events have yielded mixed-lineage genome constellations.
Some viruses with mixed-lineage genomes have become common ancestors to many later strains (i.e. the `trunk' of the phylogenetic tree) and the segment lineages involved so successful, as to become fixed within the influenza B population circulating today.

@@

\subsection*{PB1, PB2 and HA segments continue to circulate as two lineages and have not reassorted recently}
Unlike PA, NP, NA, MP and NS segments, which have periodically lost diversity, we find that PB1, PB2 and HA segments have continued to accumulate diversity since the initial split of Vic and Yam lineages (see Figure \ref{tmrcaOT}).
In addition, there is evidence to suggest that the numbers of sequences that belong to either Vic and Yam lineages of PB1, PB2 and HA segments have been sampled at a ratio close to 0.5 (see Figure \ref{lineageRatiosOverTime}), presumably due to the action of some form of balancing selection.

@@

In Figure \ref{betweenDiversity} we focus on PB1, PB2 and HA lineage labels, which have both Vic and Yam lineages up to recent times.
We find that there have been some losses of diversity, indicative of label mixing i.e. reassortment, in the early years of the Vic--Yam split (1986--1996) between HA and PB1--PB2 segments.
After 1997, however, mean pairwise between label diversity increases in PB1, PB2 and HA segments, whereby Vic and Yam labels share a common ancestor only at the root.
Our data thus suggest that since 1997 there has been no reassortments between Vic and Yam lineages of PB1, PB2 and HA segments.
In addition mean pairwise between label diversity measurements after 1997 become identical regardless of chosen label, which suggests that PB1, PB2 and HA lineage labels are assigned identically in all trees investigated.

@@

\subsection*{Recent PB1-PB2-HA constellations are entirely derived from Victoria or Yamagata lineage segments}

@@

We find that PB1, PB2 and HA segments, in addition to not reassorting across the Vic-Yam lineage boundary recently, form co-reassorting segment complexes derived entirely of Vic or Yam lineage segments.
Figure \ref{stateTime} shows the sum of branch lengths which were labelled as having entirely Vic, entirely Yam or mixed-lineage PB1, PB2 and HA segments.
Due to lack of reassortment between Vic and Yam lineages of PB1, PB2 and HA (Figure \ref{betweenDiversity}) all segments have spent significantly longer periods of time with either entirely Vic-derived or entirely Yam-derived than with mixed-lineage PB1, PB2 and HA constellations.

We have identified 3 classes of mixed-lineage PB1-PB2-HA reassortants from our data with the following PB1-PB2-HA constellations: VVY (B/Bankok/163/1990-like, 13 sequences isolated 1990 -- 5th Jan 1995), YVV (B/Nanchang/630/1994-like, 2 sequences isolated 1994 -- 1996) and YVY (B/New York/24/1993-like, 2 sequences isolated 8th Jan 1993 -- 1994).
An additional class of reassortants with a YYV PB1-PB2-HA constellation (B/Waikato/6/2005-like, 16 sequences isolated June 9 -- November 12 in 2005) was found when investigating a larger dataset with only PB1, PB2, HA and NA sequences.
In all cases PB1-PB2-HA reassortants have not persisted for prolonged periods of time and have not been fixed in the influenza B population.
In particular reassortment events combining PB1 and PB2 segments of different lineages, e.g. B/Nanchang/630/1994-like and B/New York/24/1993-like isolates, exhibit poor sampling and short circulation times.
In contrast, reassortments combining PB1+2 and HA of different lineages (B/Bankok/163/1990-like and B/Waikato/6/2005-like isolates) have more isolates and circulated for much longer periods of time in the past.

@@

@@

We reconstructed reassortment events detectable by discrete traits.
Figure \ref{railroadPlot} focuses only on reassortments that have occured after 1990.
We identify 5 major reassortant genome constellations (given in order PB1-PB2-PA-HA-NP-NA-MP-NS) circulating 1992 -- 2011: B/Alaska/12/1996-like (YYYYYYYV), B/Nanchang/2/1997-like (VVYVYVYV), B/Iowa/03/2002-like (VVYVYYYV), B/California/NHRC0001/2006-like (VVYVYYYV) and B/Brisbane/33/2008-like (VVYVYYYV).
Figure \ref{railroadPlot} shows that reassorting segments appear to evolve with a considerable degree of autonomy.
For example, the NP lineage that entered a largely Victoria lineage derived genome and gave rise to the B/Nanchang/2/1997-like isolates continued circulating until 2010, even though other segments it reassorted with in 1995 -- 1996 (PA and MP) went extinct following the reassortments that led to B/Iowa/03/2002-like isolates.
A more extreme example is the NS segment, which in B/Iowa/03/2002-like isolates (and all subsequent Vic PB1-PB2-HA isolates) has been derived from Victoria lineage that had been associated with mostly Yam lineage derived B/Alaska/12/1996-like genomes for a number of years.

\begin{figure}
	\centering		
	\includegraphics[width=0.95\textwidth]{figures/RailroadPlotDated.pdf}
	\caption{\textbf{Schematic plot of reconstructed reassortments between Victoria and Yamagata lineage segments of influenza B virus.}
Victoria and Yamagata lineage genomes are represented by 8 (red or blue, respectively) parallel lines.
Inter-lineage reassortment events are indicated by lines entering a different genome.
The angle of incoming lineages represents uncertainty in the timing of the event (mean date of the parent node and mean TMRCA of reassortant clades).
Lineage extinction dates are not shown accurately.}
	\label{railroadPlot}
\end{figure}

\subsection*{Reassortment rates are not significantly different between PB1, PB2 and HA, but reassortment distances can be}
Estimates of normalized SPR distances between pairs of segments (Figure \ref{SPRdistances}) show that there is no evidence of differences in reassortment rates between all segments, based on overlapping 95\% highest posterior density (HPD) intervals.

\begin{figure}
	\centering		
	\includegraphics[width=0.95\textwidth]{figures/InfB_normalizedApproxSPR.png}
	\caption{\textbf{Normalized approximate SPR distances between pairs of segments.}
All vertical lines indicating uncertainty are 95\% highest posterior densities (HPDs).}
	\label{SPRdistances}
\end{figure}

Unlike normalized SPR distances normalized $\Delta$TMRCA estimates between some segments (Figure \ref{deltaTMRCA}) indicate that reassortment distance (as quantified by $\Delta$TMRCA) between some segments, namely PB1-PB2-HA (but also NA-M1 and PA-M1-NS1), are much more similar to $\Delta$TMRCA estimates between replicate trees.

\begin{figure}
	\centering		
	\includegraphics[width=0.95\textwidth]{figures/InfB_normalizedMuDeltaTMRCA.png}
	\caption{\textbf{Normalized mean $\Delta$TMRCA statistics between pairs of segments.}
All vertical lines indicating uncertainty are 95\% highest posterior densities (HPDs).}
	\label{deltaTMRCA}
\end{figure}
